---
title: "MLB Payroll analysis"
author: "Lin, Po-Wei"
date: "2025-10-31"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
header-includes:
  - \usepackage{fontspec}
  - \setmainfont{Noto Serif CJK TC}
  - \setsansfont{Noto Sans CJK TC}
  - \setmonofont{JetBrains Mono}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
  )
library(tidyverse)
library(knitr)
library(ggplot2)
library(readr)
library(janitor)
library(stringr)
library(scales)
```

# Importing and Exploring the Dataset
```{r}
#Import the dataset
raw_url <- "https://raw.githubusercontent.com/pwlinbernie/MLB-payroll-analysis/refs/heads/main/mlb_payrolls.csv"
mlb <- read_csv(raw_url)

#check
dim(mlb)
head(mlb)
glimpse(mlb)
```
The dataset contains information on each MLB team’s annual payroll and performance from multiple seasons, including total payroll allocations, wins, losses, and postseason results.
Before conducting any analysis, it’s essential to understand the data structure and verify that the variables are properly imported.

# Data Cleaning and Variable Preparation
```{r}
mlb <- mlb |>
    clean_names() |>
    mutate(
      team = coalesce(team, 'team_name'),
      year = as.integer(year),
      payroll = parse_number(total_payroll_allocations),
      active26 = parse_number(active_26_man),
      wins = as.integer(wins),
      losses = as.integer(losses),
      games = wins + losses,
      win_pct = if_else(games > 0, wins / games, NA_real_),
      #postseason
      postseason =na_if(as.character(postseason), ""),
      playoffs_bin = case_when(
        is.na(postseason) ~0L,
        str_detect(str_to_lower(postseason), "no playoffs") ~0L,
        TRUE ~ 1L
      ),
      log_payroll = log(payroll)
)
# check
names(mlb)
summary(select(mlb, payroll, active26, wins, losses, win_pct, playoffs_bin))
sum(is.na(mlb$payroll)); sum(is.na(mlb$active26))
```
Standardizing column names and converts key variables into numeric formats.

payroll and active26 are parsed as numerical values.
win_pct is calculated from total wins and losses.
playoffs_bin is created as a binary indicator (1 = made playoffs, 0 = did not).
log_payroll is the natural logarithm of payroll, reducing skewness in later analysis.

# Payroll Distribution (Histogram)
```{r histogram}

```

# Payroll vs. Winning Percentage (Scatterplot with Regression Line)
```{r scatter}
ggplot(mlb, aes(x = log_payroll, y = win_pct)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
  scale_y_continuous(labels = percent) +
  labs(title = "Payroll vs. Winning Percentage",
       x = "log(Payroll)", y = "Winning Percentage") +
  facet_wrap(~ year)

```
For the scatterplot, we can see that:
1. Across almost all panels, the fitted OLS line has a positive slope, indicating that higher payroll is associated with higher winning percentage in the same season.
2. Teams with very high payrolls cluster toward the upper-right; however, there is substantial vertical spread at many payroll levels, implying that payroll alone does not fully determine performance.
3. A few low-payroll but above-.500 teams (upper-left area) and high-payroll under-.500 teams (lower-right area) illustrate that efficiency, roster construction, injuries, and coaching likely moderate the payroll–performance link.

There is a consistent positive relationship between payroll and winning percentage within seasons, but with meaningful dispersion—payroll helps, yet it does not guarantee success.

# Winning Percentage by Payroll Quartile (Boxplot)
```{r boxplot}
mlb <- mlb |>
  mutate(payroll_q = ntile(payroll, 4))

ggplot(mlb, aes(x = factor(payroll_q), y = win_pct)) +
  geom_boxplot(fill = "gray") +
  scale_y_continuous(labels = percent) +
  labs(title = "Winning Percentage by Payroll Quartile",
       x = "Payroll Quartile (1 = Lowest, 4 = Highest)",
       y = "Winning Percentage")
```
For the boxplot, we can see that:
1. The median winning percentage rises from Q1 (lowest payroll) to Q4 (highest payroll), consistent with the hypothesis that higher payrolls are associated with better performance on average.
2. Both the interquartile range (IQR) and the upper whiskers shift upward across quartiles, meaning not only the typical team but also the upper performers tend to improve with payroll.
3. Despite the upward shift, the boxes/whiskers overlap across quartiles, showing that some lower-payroll teams outperform higher-payroll teams and vice versa. Payroll is a strong correlate, not a guarantee.

The quartile comparison provides clear descriptive evidence that higher payroll groups tend to win more, yet variance and overlap indicate that other factors (roster age, injuries, development, front-office quality) remain important.

# Playoff qualification rate by payroll quartile (proportion barplot)
```{r}
mlb$Payroll_Group <- ntile(mlb$total_payroll_allocations, 4)
mlb$Playoff <- ifelse(mlb$postseason == "No Playoffs", "No", "Yes")

library(ggplot2)
library(scales)

mlb$Payroll_Group <- factor(mlb$Payroll_Group,
                            labels = c("Q1 (Lowest)", "Q2", "Q3", "Q4 (Highest)"))

ggplot(mlb, aes(x = factor(Payroll_Group), fill = Playoff)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Playoff Qualification Rate by Payroll Quartile",
       x = "Payroll Quartile",
       y = "Proportion of Teams") +
  scale_fill_manual(values = c("No" = "gray70", "Yes" = "#1f78b4"))

```
For the proportion barplot, we can see that:
1.Each bar represents a payroll quartile group (Q1–Q4), ranked from lowest to highest payroll.
2.The blue section shows the proportion of teams that qualified for the playoffs, while the gray section represents those that did not.
3.Higher-payroll teams (Q3–Q4) generally exhibit slightly higher qualification rates than lower-payroll teams (Q1–Q2).
4.However, the differences are not very large—some low-payroll teams still managed to reach the postseason.
5.This pattern indicates that payroll alone does not fully determine success; other factors such as player development, management strategy, and team efficiency also influence playoff outcomes.

In summary,while there is a positive association between team payroll and playoff qualification, the relationship is not deterministic. Payroll provides a competitive advantage, but effective management and team strategy remain key to achieving postseason success.

# Payroll vs playoff probability (loess or logistic preview)
```{r}
library(dplyr)
library(readr)   
library(ggplot2)
library(scales)

mlb <- mlb %>%
  mutate(
    payroll_num  = parse_number(total_payroll_allocations), 
    average_age  = as.numeric(average_age),
    year         = as.integer(year),
    Playoff_bin  = ifelse(postseason %in% c("No Play", "No Playoffs", "No"), 0, 1)
  )


model_logit <- glm(Playoff_bin ~ log(payroll_num) + average_age + year,
                   data = mlb, family = "binomial")


mlb$predicted_prob <- predict(model_logit, type = "response")


ggplot(mlb, aes(x = log(payroll_num), y = predicted_prob)) +
  geom_point(color = "darkred", alpha = 0.6) +
  geom_smooth(method = "loess") +
  scale_y_continuous(labels = percent, limits = c(0,1)) +
  labs(title = "Predicted Probability of Making Playoffs",
       x = "Log of Payroll", y = "Predicted Probability") +
  theme_minimal()


newdat <- with(mlb, data.frame(
  payroll_num = seq(min(payroll_num, na.rm = TRUE),
                    max(payroll_num, na.rm = TRUE), length.out = 200),
  average_age = mean(average_age, na.rm = TRUE),
  year        = max(year, na.rm = TRUE) 
))
newdat$pred <- predict(model_logit, newdata = newdat, type = "response")

ggplot() +
  geom_point(data = mlb,
             aes(x = log(payroll_num), y = Playoff_bin),
             alpha = 0.2, position = position_jitter(height = 0.02)) +
  geom_line(data = newdat,
            aes(x = log(payroll_num), y = pred), linewidth = 1) +
  scale_y_continuous(labels = percent, limits = c(0,1)) +
  labs(title = "Logistic Curve: Payroll vs. Playoff Probability",
       x = "Log of Payroll", y = "Predicted Probability") +
  theme_minimal()


```
For the playoff probability plots, we can see that:
1.The nonlinear loess smoothing plot (left) illustrates the actual data trend between team payroll and the probability of making the playoffs.
2.The blue curve shows a clear upward trend: as payroll increases, the predicted playoff probability rises noticeably, especially when the logarithm of payroll exceeds approximately 18.5.
3.The widening gray confidence bands at both ends indicate higher uncertainty due to fewer teams at extreme payroll levels.
4.The logistic regression curve (right) depicts the model-based linear-in-logit relationship between payroll and playoff probability.
5.The nearly flat black line suggests that, after controlling for year and team age, the marginal effect of payroll on playoff qualification is relatively small.
6.Comparing both plots reveals that the true relationship is nonlinear—the logistic model underestimates the sharp increase in playoff probability among top-spending teams.

In summary,payroll has a positive association with playoff qualification, but the effect is nonlinear. High-payroll teams enjoy a substantially greater likelihood of reaching the postseason, while payroll advantages become significant only beyond the top spending levels.
